CF Registry Analysis - ROC

### 2024-03-13 Christine Chan/CF Registry/cf_registry_ROC_analysis.qmd
ROC update for NACFC abstract

---
title: "CF Registry Analysis - ROC"
author: "Tim Vigers"
date: "today"
date-format: long
format:
  html:
    toc: true
    toc-depth: 5
    toc-float: true
    code-fold: true
    self-contained: true
    fig-cap-location: top
    page-layout: full
    theme:
      light: flatly
      dark: darkly
bibliography: /Users/timvigers/Dropbox/Miscellaneous/zotero.bib
csl: /Users/timvigers/Dropbox/Miscellaneous/american-medical-association.csl
editor: source
---

```{r setup}
#| include: false
library(tidyverse)
library(gtsummary)
switch(Sys.info()[["sysname"]],
  Windows = {
    home_dir <- "C:/Users/Tim/OneDrive - The University of Colorado Denver/Vigers/CF/Christine Chan/CF Registry"
  },
  Darwin = {
    home_dir <- "/Users/timvigers/Library/CloudStorage/OneDrive-TheUniversityofColoradoDenver/Vigers/CF/Christine Chan/CF Registry"
  }
)
knitr::opts_knit$set(root.dir = home_dir)
```

```{r data cleaning}
# Import data
encounter <- read.csv("./Data_Raw/CFF19_encountersMerged_Del1.csv",
  na.strings = c("", "N/A")
)
# Limit to those with two hour OGTT and Hba1c at the same encounter
roc_df <- encounter %>%
  select(eDWID, reviewyear, encounternum, twohour_bloodglucose, hgba1c) %>%
  drop_na() %>%
  arrange(eDWID, reviewyear, encounternum) %>%
  group_by(eDWID) %>%
  filter(row_number() == n()) %>%
  ungroup() %>%
  mutate(`OGTT 2-hour` = cut(twohour_bloodglucose,
    breaks = c(-Inf, 140, 200, Inf), labels = c("<140", "[140,200)", ">=200")
  ))
# Add demographic data
demo <- read.csv("./Data_Raw/CFF19_DemogCFDiag_Del1.csv",
  na.strings = c("", "N/A")
)
roc_df <- left_join(roc_df,
  demo %>% select(eDWID, Race1:Race6, Hispanicrace, Gender, MutClass),
  by = join_by(eDWID)
)
write.csv(roc_df, file = "~/ROC data.csv", row.names = F, na = "")
```

@@

Race 1: White

Race 2: Black or African American

Race 3: American Indian or Alaska Native

Race 4: Asian

Race 5: Native Hawaiian or Other Pacific Islander

Race 6: Some other race

```{r}
roc_df %>%
  select(reviewyear, hgba1c, `OGTT 2-hour`, Race1:Race6, Hispanicrace, Gender, MutClass) %>%
  tbl_summary(by = `OGTT 2-hour`, missing_text = "Missing") %>%
  add_overall()
```

@@

>>>>

---
title: "CF Registry Analysis - ROC"
author: "Tim Vigers"
date: "today"
date-format: long
format:
  html:
    toc: true
    toc-depth: 5
    toc-float: true
    code-fold: true
    self-contained: true
    fig-cap-location: top
    fig-width: 8
    fig-height: 8
    page-layout: full
bibliography: /Users/timvigers/Library/CloudStorage/OneDrive-TheUniversityofColoradoDenver/Miscellaneous/zotero.bib
csl: /Users/timvigers/Library/CloudStorage/OneDrive-TheUniversityofColoradoDenver/Miscellaneous/american-medical-association.csl
editor: source
---

```{r setup}
#| include: false
library(tidyverse)
library(gtsummary)
library(gt)
library(pROC)
switch(Sys.info()[["sysname"]],
  Windows = {
    home_dir <- "C:/Users/Tim/OneDrive - The University of Colorado Denver/Vigers/CF/Christine Chan/CF Registry"
  },
  Darwin = {
    home_dir <- "/Users/timvigers/Library/CloudStorage/OneDrive-TheUniversityofColoradoDenver/Vigers/CF/Christine Chan/CF Registry"
  }
)
knitr::opts_knit$set(root.dir = home_dir)
```

```{r data cleaning}
# Import data
encounter <- read.csv("./Data_Raw/CFF19_encountersMerged_Del1.csv",
  na.strings = c("", "N/A")
)
# Demographic data
demo <- read.csv("./Data_Raw/CFF19_DemogCFDiag_Del1.csv",
  na.strings = c("", "N/A")
)
# Race
races <- c(
  "White", "Black or African American", "American Indian or Alaska Native",
  "Asian", "Native Hawaiian or Other Pacific Islander", "Some other race"
)
demo$Race <- apply(demo, 1, function(r) {
  w <- which(r[paste0("Race", 1:6)] == 1)
  if (length(w) == 0) {
    "Unknown"
  } else if (length(w) > 1) {
    "More than one race"
  } else {
    races[w]
  }
})
demo$Race <- factor(demo$Race,
  levels = c(
    "White", "Black or African American", "American Indian or Alaska Native",
    "Asian", "More than one race", "Native Hawaiian or Other Pacific Islander",
    "Some other race"
  ),
  labels = c(
    "White", "Black or African American", "Other race", "Other race",
    "More than one race", "Other race", "Other race"
  )
)
demo$Hispanicrace <- factor(demo$Hispanicrace,
  levels = 1:2,
  labels = c("Yes", "No")
)
# F508
demo$F508 <- factor(demo$F508,
  levels = 1:3,
  labels = c("F508del/F508del", "F508del/Other", "Other/Other")
)
# Age at year end
annual <- read.csv("./Data_Raw/CFF19_Annualized_Del1.csv",
  na.strings = c("", "N/A")
)
annual <- annual %>% rename(reviewyear = ReviewYear)
# Merge
encounter <- left_join(
  encounter,
  demo %>% select(eDWID, Race, Hispanicrace, Gender, F508, MutClass),
  by = join_by(eDWID)
)
encounter <- left_join(
  encounter,
  annual %>% select(eDWID, reviewyear, Age_YrEnd),
  by = join_by(eDWID, reviewyear)
)
# Limit to those with two hour OGTT and HbA1c at the same encounter
# Also remove years before 2010 and those under age 10
roc_df <- encounter %>%
  arrange(eDWID, reviewyear, encounternum) %>%
  select(
    eDWID, reviewyear, Age_YrEnd, Race, Hispanicrace, Gender, F508, MutClass,
    twohour_bloodglucose, hgba1c
  ) %>%
  filter(reviewyear >= 2010, Age_YrEnd >= 10, hgba1c >= 3.5 & hgba1c <= 15) %>%
  drop_na() %>%
  group_by(eDWID) %>%
  filter(row_number() == n()) %>%
  ungroup() %>%
  mutate(
    `OGTT 2-hour` = cut(twohour_bloodglucose,
      breaks = c(-Inf, 140, 200, Inf), labels = c("<140", "[140,200)", ">=200")
    ),
    `Age Group` = cut(Age_YrEnd, breaks = c(10, 18, 30, Inf), right = F)
  )
# Metrics to report
metrics <- c(
  "thresh", "specificity", "sensitivity", "accuracy", "tn", "tp", "fn", "fp",
  "npv", "ppv"
)
```

# Data cleaning

1. People who reported American Indian or Alaska Native, Asian, and Native Hawaiian or Other Pacific Islander for race were considered as "Other race" due to relatively small sample sizes for these groups.

2. Years prior to 2010 and those under age 10 at the end of a given year were excluded from the analysis.

3. HbA1cs < 3.5% or > 15% were excluded from the analysis, as these values were assumed to be entered incorrectly.

@@

## By two hour OGTT value

```{r}
roc_df %>%
  select(
    reviewyear, hgba1c, Age_YrEnd, `Age Group`, `OGTT 2-hour`, Race,
    Hispanicrace, Gender, F508, MutClass
  ) %>%
  tbl_summary(
    by = `OGTT 2-hour`, missing_text = "Missing",
    type = list(reviewyear ~ "categorical"),
    label = list(
      hgba1c = "HbA1c", Hispanicrace = "Hispanic", MutClass = "Mutation Class",
      reviewyear = "Year", Age_YrEnd = "Age at Year End"
    )
  ) %>%
  add_overall() %>%
  add_p(
    test = list(reviewyear = "fisher.test"),
    test.args = all_tests("fisher.test") ~ list(simulate.p.value = T)
  ) %>%
  separate_p_footnotes() %>%
  as_gt()
```

## By age

```{r}
roc_df %>%
  select(
    reviewyear, hgba1c, Age_YrEnd, `Age Group`, `OGTT 2-hour`, Race,
    Hispanicrace, Gender, F508, MutClass
  ) %>%
  tbl_summary(
    by = `Age Group`, missing_text = "Missing",
    type = list(reviewyear ~ "categorical"),
    label = list(
      hgba1c = "HbA1c", Hispanicrace = "Hispanic", MutClass = "Mutation Class",
      reviewyear = "Year", Age_YrEnd = "Age at Year End"
    )
  ) %>%
  add_overall() %>%
  add_p(
    test = list(reviewyear = "fisher.test"),
    test.args = all_tests("fisher.test") ~ list(simulate.p.value = T)
  ) %>%
  separate_p_footnotes() %>%
  as_gt()
```

# HbA1c by diagnosis

```{r}
ggplot(roc_df, aes(x = `OGTT 2-hour`, y = hgba1c)) +
  geom_boxplot() +
  ylab("HbA1c") +
  theme_classic()
```

@@

## Three groups, two cutoffs

For the following analyses, we considered those with a two hour glucose value >= 140 and < 200 as a distinct group (AGT) and use receiver operating characteristic curves to find cutoffs that predict:

1. NGT vs. AGT

2. AGT vs. CFRD

```{r}
ngt_agt <- roc_df %>% filter(`OGTT 2-hour` != ">=200")
ngt_agt$ogtt <- droplevels(ngt_agt$`OGTT 2-hour`)
agt_cfrd <- roc_df %>% filter(`OGTT 2-hour` != "<140")
agt_cfrd$ogtt <- droplevels(agt_cfrd$`OGTT 2-hour`)
```

### Peds and adults

#### NGT vs. AGT

```{r}
#| message: false
r <- roc(ogtt ~ hgba1c, data = ngt_agt)
plot(r, print.auc = T, print.thres = T)
gt(coords(r, ret = metrics)) %>% fmt_number(n_sigfig = 3)
```

#### AGT vs. CFRD

```{r}
#| message: false
r <- roc(ogtt ~ hgba1c, data = agt_cfrd)
plot(r, print.auc = T, print.thres = T)
gt(coords(r, ret = metrics)) %>% fmt_number(n_sigfig = 3)
```

### Peds

```{r}
peds_ngt_agt <- ngt_agt %>% filter(Age_YrEnd < 18)
peds_agt_cfrd <- agt_cfrd %>% filter(Age_YrEnd < 18)
```

#### NGT vs. AGT

```{r}
#| message: false
r <- roc(ogtt ~ hgba1c, data = peds_ngt_agt)
plot(r, print.auc = T, print.thres = T)
gt(coords(r, ret = metrics)) %>% fmt_number(n_sigfig = 3)
```

#### AGT vs. CFRD

```{r}
#| message: false
r <- roc(ogtt ~ hgba1c, data = peds_agt_cfrd)
plot(r, print.auc = T, print.thres = T)
gt(coords(r, ret = metrics)) %>% fmt_number(n_sigfig = 3)
```

### Adults

```{r}
adult_ngt_agt <- ngt_agt %>% filter(Age_YrEnd >= 18)
adult_agt_cfrd <- agt_cfrd %>% filter(Age_YrEnd >= 18)
```

#### NGT vs. AGT

```{r}
#| message: false
r <- roc(ogtt ~ hgba1c, data = adult_ngt_agt)
plot(r, print.auc = T, print.thres = T)
gt(coords(r, ret = metrics)) %>% fmt_number(n_sigfig = 3)
```

#### AGT vs. CFRD

```{r}
#| message: false
r <- roc(ogtt ~ hgba1c, data = adult_agt_cfrd)
plot(r, print.auc = T, print.thres = T)
gt(coords(r, ret = metrics)) %>% fmt_number(n_sigfig = 3)
```

## Two groups, one cutoffs

For the following analyses, we examined HbA1c cutoffs that best predict a two hour OGTT value either >= 140 or >= 200

```{r}
roc_df$agt <- as.numeric(roc_df$twohour_bloodglucose >= 140)
roc_df$cfrd <- as.numeric(roc_df$twohour_bloodglucose >= 200)
```

### Peds and adults

#### $\geq$ 140

```{r}
#| message: false
r <- roc(agt ~ hgba1c, data = roc_df)
plot(r, print.auc = T, print.thres = T)
gt(coords(r, ret = metrics)) %>% fmt_number(n_sigfig = 3)
```

#### $\geq$ 200

```{r}
#| message: false
r <- roc(cfrd ~ hgba1c, data = roc_df)
plot(r, print.auc = T, print.thres = T)
gt(coords(r, ret = metrics)) %>% fmt_number(n_sigfig = 3)
```

### Peds

```{r}
peds_roc <- roc_df %>% filter(Age_YrEnd < 18)
```

#### $\geq$ 140

```{r}
#| message: false
r <- roc(agt ~ hgba1c, data = peds_roc)
plot(r, print.auc = T, print.thres = T)
gt(coords(r, ret = metrics)) %>% fmt_number(n_sigfig = 3)
```

#### $\geq$ 200

```{r}
#| message: false
r <- roc(cfrd ~ hgba1c, data = peds_roc)
plot(r, print.auc = T, print.thres = T)
gt(coords(r, ret = metrics)) %>% fmt_number(n_sigfig = 3)
```

### Adults

```{r}
adult_roc <- roc_df %>% filter(Age_YrEnd >= 18)
```

#### $\geq$ 140

```{r}
#| message: false
r <- roc(agt ~ hgba1c, data = adult_roc)
plot(r, print.auc = T, print.thres = T)
gt(coords(r, ret = metrics)) %>% fmt_number(n_sigfig = 3)
```

#### $\geq$ 200

```{r}
#| message: false
r <- roc(cfrd ~ hgba1c, data = adult_roc)
plot(r, print.auc = T, print.thres = T)
gt(coords(r, ret = metrics)) %>% fmt_number(n_sigfig = 3)
```
